A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-13 text/survival-methods.Rmd
Add all the survival code, including kmplots for simulated data, and modify some of the methods section on survival
Create a single large panel figure for the survival stuff, rewrite the text to deal with that appropriately. Also adjust the Makefile to build the script properly now that we're using .docx
Reorganize some of the survival code.

$$\frac{dP}{dT} = R(T)(1-P(T))$$

@@

The $k$ in our model can be taken as, roughly, the aging rate relative to the
baseline rates as depicted in **supplementary figure S5**. Thus, a value of
$k=2$ means that a given cancer causes you to experience the same risk in time
$dt$ as a non-cancer patient would experience at the same age, in time $2dt$.

Using this model, we split the data by Male/Female sex, and by White, vs Black
race as these were the main well-identified categories available to us in the
dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black vs White
race is an important category due to differences in access to care and other
socioeconomic and education related factors.

Using the model described above we fit the data using Stan via the rstan package
in R. Posterior inferences are shown in **FIXME WHICH FIGURE?**.

**supplementary figure S5** gives a comparison on log-scale between the post-fit
survival curves and the CDC data it is based on. For demonstration purposes we
also illustrate some examples of the expected survival for a population that
receives a diagnosis with doubled risk of death at ages 40, 50, 60, 70, and 80
(**supplementary figure S6**).

>>>>

$$\frac{dP}{dT} = (1-P(T))R(T)$$

@@

The $k$ in our model can be taken as, roughly, the aging rate relative
to the baseline rates. Thus, a value of $k=2$ means that a given
cancer causes you to experience the same risk in time $dt$ as a
non-cancer patient would experience at the same age, in time $2dt$.

Using this model, we split the data by Male/Female sex, and by White, vs Black
race as these were the main well-identified categories available to us in the
dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black vs White
race is an important category due to differences in access to care and other
socioeconomic and education related factors. However, due to the low number of
Black patient tumors, we use ethnicity only to identify the baseline risk
function, while the tissue and class specific multipliers are not ethnicity
specific.

Using the model described above we fit the data using Stan via the rstan package
in R. Posterior inferences are shown in **Figure 7**.

**supplementary figure S5** gives a comparison on log-scale between
the post-fit survival curves and the CDC data it is based on.

### 2020-08-13 text/survival.Rmd
Create a single large panel figure for the survival stuff, rewrite the text to deal with that appropriately. Also adjust the Makefile to build the script properly now that we're using .docx
Reorganize some of the survival code.

If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. Kaplan-Meier plots reveal large differences between
cancer types as expected, and some modest differences between our classes (not
shown). These analyses are profoundly limited by confounding factors of age and
stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
name the obvious ones. To explore these ideas, we tested models of survival
using public longevity data from the CDC, accounting for age, gender and
ethnicity. Our model assumes that at birth each person has a small initial
probability per unit time to die, and that this probability per unit time of
death conditional on age, grows exponentially in time at some baseline rate that
matches the observed CDC data. The rate of growth in risk is assumed to jump to
a higher level at diagnosis, which we take as a proxy for onset. This represents
an accelerated aging type model. To fit the model, we constrain the baseline
risks using the available CDC life tables. We then constrain the cancer specific
multiplier using the longevity data within the cancer tumor dataset. Our model
estimates the effect of cancer type and class-specific cancer effects
independently, resulting in a cancer and class specific estimate of the aging
rate multiplier. We found that cancer types, as expected, have a range of
prognoses relative to the general population, as shown in **figure 5a**. The
least deadly was prostate cancer (PRAD), and the most deadly was glioblastoma
(GBM) which has a risk multiplier of between 15 and 22 relative to the
background risk of death in the population. This undoubtedly is due to the fact
that relatively young patients are affected by GBM. Among the deadliest cancers
outside of GBM were stomach (STAD), melanoma (SKCM) and pancreatic (PAAD)
cancers. Apart from these top 4 cancers which ranged from 6-22 in relative risk
of age-adjusted death, the remaining cancer types ranged from about 1 to 5 in
magnitude (**figure 5a**). A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

![](../figures/fig6B.png)
![](../figures/fig6C.png)
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 5: relative survival of pathway clusters. a) Posterior distributions of estimated relative age-dependent survival by cancer type. b) Posterior distributions of estimated class-dependent survival for each cancer type. See methods for model details.</p>

Our estimates of tissue-specific class effects, in contrast, ranged from 0-3,
reflecting that some classes are either less deadly or more deadly than other
classes within each cancer types. A class specific rate of 1 represents typical
rate for this tissue type (**figure 5b**). For several cancers ( _e.g._ PRAD,
kidney chromophobe (KICH), diffuse large B-cell lymphoma (DBLC), thyroid cancer
(THCA)) the posterior estimates are largely indestinguishable from the prior,
reflecting that either there were too few mortalities in the data to make an
estimate (as expected for PRAD and THCA) or two few samples period. We did not
observe class-specific trends that held true across cancer types, which could
result from different cancers having different standards of care for example. In
support of this interpretation, we also tested a factored model which considered
both cancer type and each class independent of cancer type (**supplementary
figure S7??**). Though we were able to successfully fit this model, it is a
special case of the more general model where class and tissue can independently
affect longevity, and there is no reason to believe that class specific effects
would necessarily remain constant across tissue types given how widely the
patients vary across tissue types as well as the different drug and surgical
treatments typical used for each tissue.

```{r survival-model, eval=FALSE}

```

![**Figure 6B**](../figures/fig6B.png)

![**Figure 6C**](../figures/fig6C.png)

>>>>

If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. Kaplan-Meier plots reveal large differences between
cancer types as expected, and some modest differences between our classes (not
shown). These analyses are profoundly limited by confounding factors of age and
stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
name the obvious ones. To explore these ideas, we tested models of survival
using public longevity data from the CDC, accounting for age, gender and
ethnicity. Our model assumes that at birth each person has a small initial
probability per unit time to die, and that this probability per unit time of
death conditional on age, grows exponentially in time at some baseline rate that
matches the observed CDC data. The rate of growth in risk is assumed to jump to
a higher level at diagnosis, which we take as a proxy for onset. This represents
an accelerated aging type model. To fit the model, we constrain the baseline
risks using the available CDC life tables. We then constrain the cancer specific
multiplier using the longevity data within the cancer tumor dataset. Our model
estimates the effect of cancer type and class-specific cancer effects
independently, resulting in a cancer and class specific estimate of the aging
rate multiplier.

To understand the model, see **Figure 7A** where we show simulated survival time
curves for four related groups. We start with a cohort of women aged 30 and show
their expected survival (blue). By setting their cancer rate multiplier to k=2
we can simulate the effect of a moderately deadly cancer (red). Compare this to
a randomly selected group of women ranging in age from 30 to 70 years old (dark
blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a k=2 malignancy further reduces expected
lifespan (dark red). We can see that for different tissues simply due to the
change in distribution of age-at-diagnosis we should expect equally deadly
cancers to have *different* survival curves. The model takes this into account
and the aggressiveness of the cancer can be estimated without the confounding
aspect of age.

We found that cancer types, as expected, have a range of prognoses relative to
the general population. For example in **Figure 7B-E** we see three particularly
deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
cancer where diagnosis apparently decreases risk relative to background
(Thyroid: THCA). Cancers with posterior probability for relative risk of less
than 1 such as THCA may be those where diagnosis is associated with patients who
have relatively healthy behaviors, are seen regularly by doctors, and comply
with health recommendations.

The degree to which a given cancer accelerates aging can be determined by the
ktis multiplier (**Figure 7F**) multiplied by the class specific modifier
(**Figure 7G**). Looking at the tissue specific multiplier, the least deadly was
prostate cancer (PRAD), and the most deadly was glioblastoma (GBM) which has a
risk multiplier of between 15 and 22 relative to the background risk of death in
the population. This undoubtedly is due to the fact that relatively young
patients are affected by GBM and that it is extremely deadly. Among the
deadliest cancers outside of GBM were stomach (STAD), melanoma (SKCM) and
pancreatic (PAAD) cancers. Apart from these top 4 cancers which ranged from 6-22
in relative risk of age-adjusted death, the remaining cancer types ranged from
about 1 to 5 in magnitude. A rate of 1 represents patients who die relatively
close to the age predicted purely from their demographics independent of the
information that they have the given cancer.

Our estimates of tissue-specific class effects, in contrast, ranged from less
than 1 up to about 3 or 4, reflecting that some classes are either less deadly
or more deadly than other classes within each cancer types (**Figure 7G**). A
class specific rate of 1 represents typical rate for this tissue type. For
several cancers ( _e.g._ PRAD, kidney chromophobe (KICH), diffuse large B-cell
lymphoma (DBLC), thyroid cancer (THCA)) the posterior estimates are largely
indestinguishable from the prior, reflecting that either there were too few
mortalities in the data to make an estimate (as expected for PRAD and THCA) or
two few samples period. We did not observe class-specific trends that held true
across cancer types, which could result from different cancers having different
standards of care for example. In support of this interpretation, we also tested
a factored model which considered both cancer type and each class independent of
cancer type (not shown). Though we were able to successfully fit this model, it
is a special case of the more general model where class and tissue can
independently affect longevity, and there is no reason to believe that class
specific effects would necessarily remain constant across tissue types given how
widely the patients vary across tissue types as well as the different drug and
surgical treatments typical used for each tissue.
